Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-10-06 paper.tex
abstract, intro, concl. improved as a result of eth referee report

\begin{abstract}
  We study the possibility for detecting gamma-ray emission from galaxy
  clusters. We consider 1) cosmic ray (CR) induced pion decay, 2) different
  representative benchmark models of supersymmetric dark matter (DM), and 3)
  leptophilic models of DM annihilation that include a Sommerfeld enhancement
  (SFE). To model DM annihilation, we consider continuum emission as a result of
  the hadronization of annihilating neutralinos, internal bremsstrahlung, and
  inverse Compton emission from the cosmic microwave background as well as from
  a realistic spatial and spectral distribution of dust and stellar light. Among
  all clusters/groups, we predict Virgo, Fornax and M49 to be the brightest DM
  sources and find a particularly low CR-induced background for Fornax. For a
  minimum substructure mass given by the DM free-streaming scale, cluster halos
  maximize the substructure boost for which we find a factor of $\gtrsim 1000$.
  Since the annihilation flux of substructures is mostly contributed by the
  regions around the virial radius, the resulting surface brightness profiles
  are almost flat. This makes it very challenging to detect this flux with
  imaging atmospheric Cherenkov telescopes since their sensitivity drops
  linearly with radius and they typically have $5-10$ linear resolution elements
  across a cluster; the gamma-ray satellite \Fermi is better suited. Assuming
  cold dark matter with a substructure mass distribution down to an Earth mass
  and using \Fermi upper limits, we rule out the leptophilic models in their
  present form in 28 clusters, and limit the boost from SFE in M49 and Fornax to
  be $\lesssim 5$. This corresponds to a limit on SFE in the Milky Way of
  $\lesssim 3$. Alternatively, if the SFE is realized in Nature, this
  would imply a limiting substructure mass of $\mlim > 10^4\,\msun$---a
  challenge for structure formation. Without improved modelling of extended
  sources or stacked likelihood analyses, \Fermi will face difficulties in
  constraining our selection of DM benchmark models in the near future. The
  \Fermi upper limits are, however, closing in on our predictions for the CR
  flux using an analytic model based on cosmological hydrodynamical cluster
  simulations. We limit the CR-to-thermal pressure in nearby bright galaxy
  clusters of the \Fermi sample to $\lesssim 10\%$ and in Norma and Coma to
  $\lesssim 3\%$. Thus we will soon start to constrain the underlying CR
  physics such as shock acceleration efficiencies or CR transport properties.
\end{abstract}

@@

Dark matter (DM) has been searched for in direct detection experiments
\cite{Pato:2010zk}, at accelerators
\cite{Ellis:2001hv,Baer:2006ff,Khachatryan:2011tk} and also in
indirect detection experiments looking for signals in the cosmic-ray
spectra of antiprotons, positrons, neutrinos and all of the
electromagnetic spectrum from radio waves to gamma-rays
\cite{Bergstrom:2009ib}. So far, the improvements in direct detection
sensitivity have put this method into focus, but the situation may
change considerably the coming few years as the CERN LHC experiments
collect data, and new gamma-ray detectors are being planned, such as
the CTA \cite{Consortium:2010bc}. In fact, it has recently been
pointed out \cite{Bergstrom:2010gh} that a dedicated ground-based
gamma-ray detector would have potential that goes far beyond that of
the other methods, depending on presently unknown parameters in the
particle physics models for DM.

@@

1. The galactic center (g.c.). This is where all numerical simulations
of cold dark matter (CDM) predict the highest density. However, the
detailed DM density in the very central part is difficult to predict,
due a to a possibly very complicated interplay between baryons, DM,
and the central galactic black hole. Also, it is a very crowded region
with many gamma-ray sources like pulsars and other supernova remnants,
which have to be subtracted from the data to extract the DM induced
signal. In fact, there is a recent claim of an indication of a
relatively light DM particle contribution to the gamma-ray flux from
the g.c. \cite{2010arXiv1010.2752H},
but other hypotheses seem to work
at least as well \cite{2010arXiv1012.5839B}.

2. The dwarf spheroidal galaxies orbiting the Milky Way (MW), like
Segue-1, Ursa Minor, Draco, Sagittarius, Sculptor, Carina or Willman-1
\cite{2009JCAP...01..016B,2010ApJ...720.1174A,2010JCAP...01..031S,2010JCAP...01..031S,2011arXiv1103.0477T,2011APh....34..608H}. The
problem here is that the nature of many of these small, dark
matter-dominated galaxies is not entirely clear, and the velocity
dispersion estimates are based on rather small numbers of stars. Tidal
disruption and confusion with star clusters are other
complications. Thus the DM density profile is very uncertain for most
of them. Nonetheless, by stacking the data together from many dwarf
spheroidals these uncertainties can be made less severe, and
preliminary results from \Fermi-LAT shows this method to give quite
promising results \cite{garde}.

3. Galaxy clusters. This possibility has been less studied
theoretically, however currently there is an ongoing observational
campaign to detect gamma-ray emission from galaxy clusters
\cite{2010ApJ...710..634A,2010JCAP...05..025A,2009A&A...495...27A,2009IJMPD..18.1627D,2009A&A...502..437A,2009arXiv0907.5000G,2009ApJ...704..240K,2009ApJ...706L.275A,2010ApJ...717L..71A}. In
fact, we noted in a previous Letter \cite{2009PhRvL.103r1302P} that
there are certain advantages that work in favour of this possible
target for gamma-ray detection of DM annihilation. Galaxy clusters
constitute the most massive objects in our Universe that are forming
today. This causes their DM subhalo mass function to be less affected
by tidal stripping compared to galaxy sized halos that formed long
ago. The annihilation luminosities of the DM halo component for
e.g. the Virgo cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in favour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions. According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos. For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than for galactic haloes.\footnote{All halo masses and length scales
  in this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define
  the virial mass $\mvir$ and virial radius $\rvir$ as the mass and
  radius of a sphere enclosing a mean density that is 200 times the
  critical density of the Universe $\rho_{\rmn{cr}}$.}

For a satellite dwarf galaxy, however, once it is accreted by the MW,
the outer regions are severely affected by tidal stripping. The longer
a satellite has been part of our Galaxy, and the closer it comes to
the center during its pericentral passage, the more material is
removed \cite{2004MNRAS.355..819G}.

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize in
Table~\ref{tab:flux_tab} our results for the gamma-ray flux from four
promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. However, if the boost from Sommerfeld
enhancement (SFE) is small then the gamma-ray emission induced by CRs
is dominating most clusters with the exception of a few clusters; most
notably Fornax that shows the largest ratio of DM annihilation-to-CR
flux. In particular, we identify Fornax and Virgo as prime targets for
DM observations, Perseus for CR induced emission and we additionally
decide in favor of the well studied cluster Coma for comparison. While
studying these four clusters in great detail, we complement the
analysis by calculating the expected gamma-ray fluxes of a nearby
X-ray flux complete sample of galaxy clusters from the Rosat all-sky
survey (extended HIghest X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
\cite{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
  we have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.} For
previous work related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
and for recent work on CRs in clusters see
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

This work studies the prospects for detecting extended gamma-ray
emission from galaxy clusters induced by both DM as well as CR
protons. In Section 2 we discuss the theory of DM and gamma-rays. In
particular, we focus on the leptophilic (LP) and supersymmetric
benchmark (BM) DM models in this work, and outline the framework
for estimating the gamma-ray emission from various radiative
processes. In Section 3 we calculate the gamma-ray fluxes and spectral
distributions from four clusters and compare it to upper limits set by
\Fermi-LAT. For the same clusters we derive the gamma-ray surface
brightness profiles in Section 4. To extend the analysis to more
clusters and find promising targets we estimate the fluxes from DM and
CRs of all clusters in the extended HIFLUGCS sample in Section 5. In
Section 6 we conclude and discuss our results.

@@

We continue by comparing the estimated differential flux from Fornax
to three other clusters in Fig.~\ref{fig:clu_comp}; the close-by and
well studied Virgo cluster, the X-ray bright Perseus clusters, and the
massive merging Coma cluster. We find high DM induced gamma-ray fluxes
from the Fornax and Virgo clusters, which confirms them as promising
targets for indirect DM searches. Especially in the $1-10$~GeV energy
range it is quite striking how constraining the upper limits are for
some of the clusters. At these energies, \Fermi-LAT has its peak
sensitivity due to a combination of increasing effective area and
decreasing source spectra as a function of energy. The upper limits
for Virgo and Perseus are unfortunately background contaminated by AGN
activity from M87 and NGC 1267, respectively, and do not gain much
from the increased sensitivity. In addition, the angular radius of
Virgo is about $6^\circ$, while the extended upper limit from
\Fermi-LAT is calculated assuming a $\sigma=1.2^\circ$ radius. In
fact, the ratio of the virial radius to the assumed extension for the
upper limits $\rvir/\sigma$ for $($Fornax, Coma, Virgo$)$ are
$(3.7,\,1.6,\,5.3)$, respectively. Similarly, it is interesting to
compare the radius that contains 68\% of the flux to the assumed
extension, $r_{68}/\sigma$, which is given by $(2.3,\,1.0,\,3.3)$ for
$($Fornax, Coma, Virgo$)$, respectively. This further motivates a
re-calculation of the \Fermi upper limits to account for the full
extension of the galaxy clusters. We also find that the theoretical
expectation of a high CR-induced flux in Coma and Perseus may
significantly complicate indirect DM searches in those clusters if the
boost factors are much lower than assumed in this work. The Fornax
cluster, however, is a great target for indirect DM studies because
the relative low gamma-ray flux from CRs, absents of an active AGN,
and high DM gamma-ray flux.

@@

{\em Dark matter annihilation.} Once supersymmetric cold dark matter
decouples from the expanding universe it streams and erases any
potentially present smaller scales. When gravitational instability
causes halos to collapse, this free-streaming scale imprints as a
characteristic minimum halo mass $M_\mathrm{min}$ into the hierarchy
of structure formation and should still survive until the present time
as substructure within larger halos. We show that the boost due to
substructures dominates over the smooth component for halo masses
$\mvir>10^3 \msun$ and contributes a constant luminosity for each
decade in halo mass. The ratio of halo-to-minimum subhalo mass
$\mvir/M_\mathrm{min}$ is maximized for the largest virialized systems
in the Universe---galaxy clusters. The corresponding substructure
boost can reach values exceeding $10^3$ which should make nearby
clusters the brightest DM annihilation sources after the MW center and
hypothetical very close-by DM substructures. We stress that this
conclusion relies on two hypotheses that are still open problems in
particle physics and numerical cosmology. We assume (i) the existence
of cold DM, i.e., the density fluctuation power spectrum extends down
to the DM free streaming scale of about an Earth mass, and (ii) assume
that structure formation at scales smaller than currently resolved
ones ($\sim 10^5\,\msun$) proceeds scale-invariant, i.e., there is no
non-linear mode coupling between different scales. This could
potentially erase structures smaller than a characteristic one as the
linear power spectrum approaches $\Delta^2(k) \propto k^3
P(k)=\rmn{const.}$ implying that structure on different scales may not
form any more hierarchically, but may collapse at the same time.

Since the mass density of substructures peaks at radii close to
$\rvir$, the contribution to the annihilation emission from
substructures is also dominated from these outer cluster regions.
This implies an almost flat surface brightness profile of the
annihilation emission which makes it necessary to have the entire
cluster in the field-of-view to take advantage of the total
substructure boost. This property makes it difficult to detect DM
annihilation emission without an enhancement of the particle physics
cross section over its standard value of $\sigma v\sim 3\times
10^{-26} ~\mathrm{cm}^3~\mathrm{s}^{-1}$ with imaging air Cherenkov
telescopes due to the spatially extended boost from substructures of
angular extent $\sim 0.5-1\deg$ (as independently found by reference
\cite{2011arXiv1104.3530S}). In contrast, the sensitivity of Cherenkov
telescopes drops linearly with radius outside the point spread
function $\sim 0.1\deg$, unless new background subtraction schemes can
be found, e.g., for the survey mode of CTA.

@@

We identify Virgo, Fornax and M49 to be the best cluster/group
candidates for indirect DM studies as they are expected to emit the
brightest annihilation flux.\footnote{We excluded Virgo from our
  further analysis due to its large angular extent that needs a
  separate treatment.} More importantly, Fornax has a comparably low
CR induced gamma-ray flux which may enable an indirect DM detection or
a particularly tight limit on DM properties. Other nearby bright
sources of the extended HIFLUGCS sample of nearby bright X-ray
cluster/groups are Ophiuchus, and Centaurus.

The non-detection of gamma-ray emission by \Fermi in a total of 28
clusters/groups considerably constrains LP models for DM annihilation
which were introduced to explain the increasing positron-to-lepton
ratio beyond 10 GeV. Assuming that the minimum DM substructure mass
extends down to an Earth mass and that the DM annihilation flux is
dominated by small scale substructure, we limit the saturated
Sommerfeld enhancement factor in M49 and Fornax to $\lesssim 5$. This
corresponds to a value in the Milky Way (MW) of $\lesssim3$, due to
the larger Sommerfeld boost factor in smaller mass halos which have a
lower velocity dispersion that enhances the particle cross section
(while assuming universality of the DM model). This would rule out LP
models in their current form based on the non-observation of
gamma-rays in any of the fore-mentioned
clusters/groups. Alternatively, assuming the LP models to be correct,
we limit the minimum substructure mass to $>10^4~\msun$ in M49 and
Fornax.

{\em Cosmic ray induced emission.} We substantially improve the
modelling of the expected gamma-ray signal from pion-decay resulting
from hadronic CR interactions with gas protons over previous work
\cite{2010ApJ...717L..71A}. We employ the analytic model of the CR
spectrum and spatial distribution that is based on cosmological
hydrodynamical simulations of galaxy clusters that self-consistently
follow the evolution of CRs in cosmic structure formation
\cite{2010MNRAS.409..449P}. Adopting the model for all clusters in the
HIFLUGCS sample of flux-limited X-ray clusters, we compare the
expected gamma-ray emission to that inferred from using
$L_\gamma$-cluster mass scaling relation. We note that the predictions
of the ``optimistic'' model in \cite{2010MNRAS.409..449P} are in
tension with \Fermi upper limits for Norma, Coma, and Ophiuchus. In
the next 2-3 years \Fermi will be able to probe the predictions of
their ``conservative'' CR model. This will enable to put realistic
limits on a combination of the acceleration efficiency of CRs in
structure formation shocks and CR transport coefficients in clusters.

>>>>

\begin{abstract}
  We study the possibility for detecting gamma-ray emission from galaxy
  clusters. We consider 1) leptophilic models of dark matter (DM) annihilation
  that include a Sommerfeld enhancement (SFE), 2) different representative
  benchmark models of supersymmetric DM, and 3) cosmic ray (CR) induced pion
  decay. To model DM annihilation, we consider continuum emission as a result of
  the hadronization of annihilating neutralinos, internal bremsstrahlung, and
  inverse Compton emission from the cosmic microwave background as well as from
  a realistic spatial and spectral distribution of dust and stellar light. Among
  all clusters/groups of a flux-limited X-ray sample, we predict Virgo, Fornax
  and M49 to be the brightest DM sources and find a particularly low CR-induced
  background for Fornax. For a minimum substructure mass given by the DM
  free-streaming scale, cluster halos maximize the substructure boost for which
  we find a factor of $\gtrsim 1000$. Since regions around the virial radius
  dominate the annihilation flux of substructures, the resulting surface
  brightness profiles are almost flat. This makes it very challenging to detect
  this flux with imaging atmospheric Cherenkov telescopes since their
  sensitivity drops approximately linearly with radius and they typically have
  $5-10$ linear resolution elements across a cluster. Assuming cold dark matter
  with a substructure mass distribution down to an Earth mass and using extended
  \Fermi upper limits, we rule out the leptophilic models in their present form
  in 28 clusters, and limit the boost from SFE in M49 and Fornax to be $\lesssim
  5$. This corresponds to a limit on SFE in the Milky Way of $\lesssim 3$ which
  is too small to account for the increasing positron fraction with energy as
  seen by PAMELA. Alternatively, if SFE is realized in Nature, this would imply
  a limiting substructure mass of $\mlim > 10^4\,\msun$---a problem for
  structure formation in most particle physics models. Using individual cluster
  observations, it will be challenging for \Fermi to constrain our selection of
  DM benchmark models without SFE. The \Fermi upper limits are, however, closing
  in on our predictions for the CR flux using an analytic model based on
  cosmological hydrodynamical cluster simulations. We limit the CR-to-thermal
  pressure in nearby bright galaxy clusters of the \Fermi sample to $\lesssim
  10\%$ and in Norma and Coma to $\lesssim 3\%$. Thus we will soon start to
  constrain the underlying CR physics such as shock acceleration efficiencies or
  CR transport properties.
\end{abstract}

@@

Dark matter (DM) has been searched for in direct detection experiments
\cite{Pato:2010zk}, at accelerators
\cite{Ellis:2001hv,Baer:2006ff,Khachatryan:2011tk} and also in
indirect detection experiments looking for signals in the cosmic-ray
(CR) spectra of antiprotons, positrons, neutrinos and all of the
electromagnetic spectrum from radio waves to gamma-rays
\cite{Bergstrom:2009ib}. So far, the improvements in direct detection
sensitivity have put this method into focus, but the situation may
change considerably the coming few years as the CERN LHC experiments
collect data, and new gamma-ray detectors are being planned, such as
the CTA \cite{Consortium:2010bc}. In fact, it has recently been
pointed out \cite{Bergstrom:2010gh} that a dedicated ground-based
gamma-ray detector would have potential that goes far beyond that of
the other methods, depending on presently unknown parameters in the
particle physics models for DM.

@@

1. The galactic center (g.c.). This is where all numerical simulations of cold
dark matter (CDM) predict the highest density. However, the detailed DM density
in the very central part is difficult to predict, due a to a possibly very
complicated interplay between baryons, DM, and the central galactic black
hole. Also, it is a very crowded region with many gamma-ray sources like
pulsars, CR-illuminated molecular clouds, and other supernova remnants, which
have to be subtracted from the data to extract the DM induced signal. In fact,
there is a recent claim of an indication of a relatively light DM particle
contribution to the gamma-ray flux from the g.c. \cite{2010arXiv1010.2752H}, but
other hypotheses seem to work at least as well \cite{2010arXiv1012.5839B}.

2. The dwarf spheroidal galaxies orbiting the Milky Way (MW), like Segue-1, Ursa
Minor, Draco, Sagittarius, Sculptor, Carina or Willman-1
\cite{2009JCAP...01..016B,2010ApJ...720.1174A,2010JCAP...01..031S,2010JCAP...01..031S,2011arXiv1103.0477T,2011APh....34..608H}. The
problem here is that the nature of many of these small, dark matter-dominated
galaxies is not entirely clear, and the velocity dispersion estimates are based
on rather small numbers of stars.  Confusion with star clusters and tidal
disruption are other complications. Once a satellite dwarf galaxy is accreted by
the MW, the outer regions are severely affected by tidal stripping. The longer a
satellite has been part of our Galaxy, and the closer it comes to the center
during its pericentral passage, the more material is removed
\cite{2004MNRAS.355..819G}.  Thus the DM density profile is very uncertain for
most of them, especially for radii larger than those probed by the
stars. Nonetheless, by stacking the data together from many dwarf spheroidals
these uncertainties can be made less severe, and preliminary results from
\Fermi-LAT shows this method to give quite promising results \cite{garde}.

%For a satellite dwarf galaxy, however, once it is accreted by the MW,
%the outer regions are severely affected by tidal stripping. The longer
%a satellite has been part of our Galaxy, and the closer it comes to
%the center during its pericentral passage, the more material is
%removed \cite{2004MNRAS.355..819G}.

3. Galaxy clusters. This possibility has been less studied
theoretically, however currently there is an ongoing observational
campaign to detect gamma-ray emission from galaxy clusters
\cite{2003ApJ...588..155R,2006ApJ...644..148P,2008AIPC.1085..569P,2009A&A...495...27A,2009arXiv0907.0727T,2009arXiv0907.5000G,2009IJMPD..18.1627D,2009A&A...495...27A,2009A&A...502..437A,2009ApJ...704..240K,2009ApJ...706L.275A,2010ApJ...710..634A,2010JCAP...05..025A,2010ApJ...717L..71A}.
In fact, we noted in a previous Letter \cite{2009PhRvL.103r1302P} that
there are certain advantages that work in favour of this possible
target for gamma-ray detection of DM annihilation. Galaxy clusters
constitute the most massive objects in our Universe that are forming
today. This causes their DM subhalo mass function to be less affected
by tidal stripping compared to galaxy sized halos that formed long
ago. The annihilation luminosities of the DM halo component for
e.g. the Virgo cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in favour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions. According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos. For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than for galactic haloes.\footnote{All halo masses and length scales
  in this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define
  the virial mass $\mvir$ and virial radius $\rvir$ as the mass and
  radius of a sphere enclosing a mean density that is 200 times the
  critical density of the Universe $\rho_{\rmn{cr}}$.}

In this paper, we will investigate in detail the potential of several
of the most promising galaxy clusters to produce an annihilation
gamma-ray yield which could be observable with present and planned
gamma-ray detectors. Here, we sketch out the main arguments that need
to be considered for maximizing the expected signal-to-noise ratio of
a promising target cluster. First, we need to maximize the DM
annihilation flux of an unresolved cluster, $F=A_\rmn{dm} \int dV
\rho^2 / D^2 \sim \mvir^\alpha/D^2$, where $A_\rmn{dm}$ depends on the
particle physics model of DM, $\rho$ is the smooth DM density profile,
$D$ is the luminosity distance, and $\alpha=1$ if we assume
universality of the DM density profile. The dependence of $\rho$ on
the halo formation epoch breaks the universality and slightly modifies
the mass dependence, yielding $\alpha=0.83$
\cite{2009PhRvL.103r1302P}. Additionally, the presence of a hierarchy
of substructures down to small scales \cite{2008Natur.456...73S} and
the potential dependence of the particle physics cross section on the
relative DM particle velocities \cite{ArkaniHamed:2008qn} may
furthermore modify the scaling parameter $\alpha$ and shall be one of
the focus points of this work. Second, we need to minimize the
expected noise which is a sum of galactic and cluster-intrinsic
foreground. While the galactic foreground varies across the sky, it is
typically lower for increasing latitude (away from the galactic
plane). It is thought that the gamma-ray foreground from clusters is
dominated by CRs which are accelerated at cosmological formation
shocks and transported over time although there could be a substantial
contribution of CRs from AGNs and supernova-driven galactic winds. In
this work we will also study (and constrain) the CR-induced emission
from clusters to identify those objects that are expected to be
especially dim which would imply a low CR-induced background.  Since
the CR-induced emission is expected to scale with the thermal X-ray
emission of clusters \cite{2004A&A...413...17P}, we estimate gamma-ray
fluxes of a nearby X-ray flux complete sample of galaxy clusters from
the Rosat all-sky survey (extended HIghest X-ray FLUx Galaxy Cluster
Sample, HIFLUGCS,
\cite{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that we
  have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.}  This
complements and extends previous work related to DM in clusters
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C}
and to CRs in clusters
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

In Section 2 we discuss the theory of DM and gamma-rays. In
particular, we focus on the leptophilic (LP) and supersymmetric
benchmark (BM) DM models in this work, and outline the framework for
estimating the gamma-ray emission from various radiative processes. In
Section 3 we calculate the gamma-ray fluxes and spectral distributions
of four clusters that are identified as prime targets for DM
observations (Fornax and Virgo) as well as to have a high CR-induced
gamma-ray yield (Perseus and Coma). For the same clusters we derive
the gamma-ray surface brightness profiles in Section 4. To extend the
analysis to more clusters and increase the list of promising targets we
estimate the fluxes from DM and CRs of all clusters in the extended
HIFLUGCS sample in Section 5. In Section 6 we conclude and discuss our
results.

% In this paper, we will investigate in some more detail the potential
% of several of the most promising galaxy clusters to produce an
% annihilation gamma-ray yield which could be observable with present
% and planned gamma-ray detectors. For convenience, we summarize in
% Table~\ref{tab:flux_tab} our results for the gamma-ray flux from four
% promising clusters with a high gamma-ray yield. We will in the
% following derive the emission from two different DM models and a
% hadronic CR model to show that Virgo has the highest gamma-ray flux
% from annihilating DM. However, if the boost from Sommerfeld
% enhancement (SFE) is small then the gamma-ray emission induced by CRs
% is dominating most clusters with the exception of a few clusters; most
% notably Fornax that shows the largest ratio of DM annihilation-to-CR
% flux. In particular, we identify Fornax and Virgo as prime targets for
% DM observations, Perseus for CR induced emission and we additionally
% decide in favor of the well studied cluster Coma for comparison. While
% studying these four clusters in great detail, we complement the
% analysis by calculating the expected gamma-ray fluxes of a nearby
% X-ray flux complete sample of galaxy clusters from the Rosat all-sky
% survey (extended HIghest X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
% \cite{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
%   we have added the Virgo cluster to the sample that we nevertheless
%   refer to as extended HIFLUGCS catalogue in the following.} For
% previous work related to DM in clusters, see, e.g.,
% \cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
% and for recent work on CRs in clusters see
% \cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

@@

We continue by comparing the estimated differential flux from Fornax
to three other clusters in Fig.~\ref{fig:clu_comp}; the close-by and
well studied Virgo cluster, the X-ray bright Perseus clusters, and the
massive merging Coma cluster. We summarize in Table~\ref{tab:flux_tab}
the expected gamma-ray flux in our DM and CR models and the
corresponding boost factors.  We find high DM induced gamma-ray fluxes
from the Fornax and Virgo clusters, which confirms them as promising
targets for indirect DM searches. Especially in the $1-10$~GeV energy
range it is quite striking how constraining the upper limits are for
some of the clusters. At these energies, \Fermi-LAT has its peak
sensitivity due to a combination of increasing effective area and
decreasing source spectra as a function of energy. The upper limits
for Virgo and Perseus are unfortunately background contaminated by AGN
activity from M87 and NGC 1275, respectively, and do not gain much
from the increased sensitivity. In addition, the angular radius of
Virgo is about $6^\circ$, while the extended upper limit from
\Fermi-LAT is calculated assuming a $\sigma=1.2^\circ$ radius. In
fact, the ratio of the virial radius to the assumed extension for the
upper limits $\rvir/\sigma$ for $($Fornax, Coma, Virgo$)$ are
$(3.7,\,1.6,\,5.3)$, respectively. Similarly, it is interesting to
compare the radius that contains 68\% of the flux to the assumed
extension, $r_{68}/\sigma$, which is given by $(2.3,\,1.0,\,3.3)$ for
$($Fornax, Coma, Virgo$)$, respectively. This further motivates a
re-calculation of the \Fermi upper limits to account for the full
extension of the galaxy clusters. We also find that the theoretical
expectation of a high CR-induced flux in Coma and Perseus may
significantly complicate indirect DM searches in those clusters if the
boost factors are much lower than assumed in this work. The Fornax
cluster, however, is a great target for indirect DM studies because
the relative low gamma-ray flux from CRs, absents of an active AGN,
and high DM gamma-ray flux.

@@

{\em Dark matter annihilation.} Once supersymmetric cold DM decouples from the
expanding universe it streams and erases any potentially present smaller
scales. When gravitational instability causes halos to collapse, this
free-streaming scale imprints as a characteristic minimum halo mass
$M_\mathrm{min}$ into the hierarchy of structure formation and should still
survive until the present time as a mass cutoff in the substructure mass
function within larger halos. We show that the boost due to substructures
dominates over the smooth component for halo masses $\mvir>10^3 \msun$ and
contributes a constant luminosity for each decade in subhalo mass. The ratio of
halo-to-minimum subhalo mass $\mvir/M_\mathrm{min}$ is maximized for the largest
virialized systems in the Universe---galaxy clusters. The corresponding
substructure boost can reach values exceeding $10^3$ which should make nearby
clusters the brightest DM annihilation sources after the MW center and
hypothetical very close-by DM substructures. We stress that this conclusion
relies on two hypotheses that are still open problems in particle physics and
numerical cosmology. We assume (i) the existence of cold DM, i.e., the density
fluctuation power spectrum extends down to the DM free streaming scale of about
an Earth mass, and (ii) assume that structure formation at scales smaller than
currently resolved ones ($\sim 10^5\,\msun$) proceeds scale-invariant, i.e.,
there is no non-linear mode coupling between different scales. This could
potentially erase structures smaller than a characteristic one as the linear
power spectrum approaches $\Delta^2(k) \propto k^3 P(k)=\rmn{const.}$ implying
that structure on different scales may not form any more hierarchically, but may
collapse at the same time.

Since the mass density of substructures peaks at radii close to $\rvir$, the
contribution to the annihilation emission from substructures is also dominated
from these outer cluster regions.  This implies an almost flat surface
brightness profile of the annihilation emission which makes it necessary to have
the entire cluster in the field-of-view to take advantage of the total
substructure boost. This property makes it difficult to detect DM annihilation
emission without an enhancement of the particle physics cross section over its
standard value of $\sigma v\sim 3\times 10^{-26} ~\mathrm{cm}^3~\mathrm{s}^{-1}$
with imaging air Cherenkov telescopes due to the spatially extended boost from
substructures over an angular extent of $\sim 0.5-1\deg$ (as independently found
by reference \cite{2011arXiv1104.3530S}). This large angular extent is not well
matched with the sensitivity of Cherenkov telescopes which drops approximately
linearly with radius outside the point spread function (with a characteristic
scale of $\sim 0.1\deg$), unless new background subtraction schemes can be
found, e.g., for the survey mode of CTA.

@@

We identify Virgo, Fornax and M49 to be the best cluster/group candidates for
indirect DM studies as they are expected to emit the brightest annihilation
flux.\footnote{We excluded Virgo from our further analysis due to its large
  angular extent that needs a separate treatment.} More importantly, Fornax has
a comparably low CR-induced gamma-ray flux which may enable an indirect DM
detection or a particularly tight limit on DM properties. Other nearby bright
sources of the extended HIFLUGCS sample are Ophiuchus, and Centaurus.

The non-detection of gamma-ray emission by \Fermi in a total of 28
clusters/groups considerably constrains LP models for DM annihilation which were
introduced to explain the increasing positron-to-lepton ratio beyond 10
GeV. Assuming that the minimum DM substructure mass extends down to an Earth
mass and that the DM annihilation flux is dominated by small scale substructure,
we limit the saturated Sommerfeld enhancement factor in M49 and Fornax to
$\lesssim 5$. This corresponds to a value in the Milky Way (MW) of $\lesssim3$,
due to the larger Sommerfeld boost factor in smaller mass halos which have a
lower velocity dispersion that enhances the particle cross section (while
assuming universality of the DM model). This would rule out LP models in their
current form based on the non-observation of gamma-rays in any of the
fore-mentioned clusters/groups and hence strongly challenge the DM
interpretation of the increasing positron fraction with energy as seen by
PAMELA. Alternatively, assuming the LP models to be correct, this would limit
the minimum substructure mass to $>10^4~\msun$ in M49 and Fornax which presents
a problem for structure formation in most particle physics models
\cite{2009NJPh...11j5027B}.

{\em Cosmic ray-induced emission.} We substantially improve the modelling of the
expected gamma-ray signal from pion-decay resulting from hadronic CR
interactions with gas protons over previous work \cite{2010ApJ...717L..71A}. We
employ the analytic model of the CR spectrum and spatial distribution that is
based on cosmological hydrodynamical simulations of galaxy clusters that
self-consistently follow the evolution of CRs in cosmic structure formation
\cite{2010MNRAS.409..449P}. Adopting the model for all clusters in the HIFLUGCS
sample of flux-limited X-ray clusters, we compare the expected gamma-ray
emission to that inferred from using an $L_\gamma-\mvir$ scaling relation. We note
that the predictions of the ``optimistic'' model in \cite{2010MNRAS.409..449P}
are in tension with \Fermi upper limits for Norma, Coma, and Ophiuchus. In the
next 2-3 years \Fermi will be able to probe the predictions of their
``conservative'' CR model. This will enable us to put realistic limits on a
combination of the acceleration efficiency of CRs in structure formation shocks
and CR transport coefficients in clusters.
